---
title: "Cumulative water deficit and surplus example"
author: "Beni Stocker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cumulative water deficit and surplus example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
```

This demonstrates the workflow for determining cumulative water deficit (CWD) time series and fitting an extreme value distribution to annual maxima of the CWD time series.

## Prepare data

Read data from the file contained in this repository.
```{r}
df <- readRDS(file = paste0(here(), "/data/df_US-Var.rds")) |> 
  # convert to from kPA to Pa (SI units are used for inputs to functions in the package)
  mutate(PA_F = 1e3 * PA_F) |> 
  
  filter(year(TIMESTAMP) >= 2001)
```

## Convert ET to mass units

Convert latent heat flux (W/m2) to evapotranspiration in mass units (mm/d).
```{r}
# tested: identical results are obtained with:
# bigleaf::LE.to.ET(LE_F_MDS, TA_F_MDS)* 60 * 60 * 24
le_to_et <- function(le, tc, patm){
  1000 * 60 * 60 * 24 * le / (cwd::calc_enthalpy_vap(tc) * cwd::calc_density_h2o(tc, patm))
}

df <- df |>
  mutate(et = le_to_et(LE_F_MDS, TA_F_MDS, PA_F))
```

Check annual totals.
```{r}
adf <- df |>
  mutate(year = year(TIMESTAMP)) |>
  group_by(year) |>
  summarise(et = sum(et), prec = sum(P_F))

adf |>
  tidyr::pivot_longer(cols = c(et, prec), names_to = "Flux") |> 
  ggplot(aes(x = year, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)")
```

Each year, the annual precipitation is greater than ET. Hence, the water deficit will not continue accumulating over multiple years.

## Simulate snow

Simulate snow accumulation and melt based on temperature and precipitation.
```{r}
df <- df |>
  mutate(prec = ifelse(TA_F_MDS < 0, 0, P_F),
         snow = ifelse(TA_F_MDS < 0, P_F, 0)) |>
  cwd::simulate_snow(varnam_prec = "prec", varnam_snow = "snow", varnam_temp = "TA_F_MDS")
```

Visualise snow mass equivalent time series.
```{r}
df |> 
  ggplot(aes(TIMESTAMP, snow_pool)) + 
  geom_line() +
  labs(x = "Date", x = "Snow mass equivalent (mm)")
```

This looks like it's a lot of snow, actually. Maybe the melting rate is too slow.

Define the daily water balance as liquid water infiltrating into soil (taken as rain plus snow melt) minus evapotranspiration - both in mass units, or equivalently in mm/d.
```{r}
df <- df |>
  mutate(wbal = liquid_to_soil - et)
```

Visualise it.
```{r}
gg5 <- df |>
  ggplot(aes(TIMESTAMP, et)) +
  geom_line()

gg6 <- df |>
  ggplot(aes(TIMESTAMP, liquid_to_soil)) +
  geom_line()

gg7 <- df |>
  ggplot(aes(TIMESTAMP, wbal)) +
  geom_line()

gg5 / gg6 / gg7
```

## Cumulative water deficit algorithm

Get CWD and events.
```{r}
out_cwd <- cwd(
  df,
  varname_wbal = "wbal",
  varname_date = "TIMESTAMP",
  thresh_drop = 0.0,
  do_surplus = TRUE
  )
```

Some preparations
```{r}
# get list of largest annual deficit events
inst_ann <- out_cwd$inst |>
  mutate(year = lubridate::year(date_start)) |>
  group_by(year) |>
  filter(max_deficit == max(max_deficit, na.rm = TRUE)) |>
  ungroup()
  
# add date of maximum defict and surplus
inst_ann <- inst_ann |> 
  mutate(date_max_deficit = out_cwd$df$TIMESTAMP[inst_ann$idx_max_deficit])
```


Plot cumulative water deficit and surplus time series.
```{r}
ggplot() +
  geom_rect(
    data = inst_ann,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_vline(xintercept = inst_ann$date_max_deficit) +
  geom_line(data = out_cwd$df, aes(TIMESTAMP, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 170)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water deficit (mm)")

ggplot() +
  geom_rect(
    data = out_cwd$inst_surplus,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_vline(xintercept = out_cwd$inst_surplus$date_start) +
  geom_line(data = out_cwd$df, aes(TIMESTAMP, surplus), color = "royalblue") +
  coord_cartesian(ylim = c(0, 1000)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water surplus (mm)")
```

Match annual maximum deficits with preceding annual maximum surpluses.
```{r}
out_cwd$inst_surplus <- out_cwd$inst_surplus |> 
  mutate(date_max_surplus = out_cwd$df$TIMESTAMP[out_cwd$inst_surplus$idx_max_surplus])
  
# merge the two data frames so that the date in inst_ann is aligned with the next 
# earlier date in out_cwd$inst_surplus
df_inst_combined <- inst_ann %>%
  left_join(
    out_cwd$inst_surplus %>% 
      select(date_max_surplus, max_surplus) |> 
      rename(date_max_preceding_surplus = date_max_surplus),
    join_by(date_max_deficit >= date_max_preceding_surplus)
  ) %>%
  group_by(date_max_deficit) %>%
  slice_max(date_max_preceding_surplus, n = 1, with_ties = FALSE) %>%
  ungroup()
```

Are the magnitudes of deficits and preceding surpluses correlated?
```{r}
df_inst_combined |> 
  ggplot(aes(max_surplus, max_deficit)) +
  geom_point()
```

No.

