---
title: "Example"
author: "Beni Stocker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{map2tidy functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
```

This demonstrates the workflow for determining cumulative water deficit (CWD) time series and fitting an extreme value distribution to annual maxima of the CWD time series.

## Prepare data

Read data from the file contained in this repository.
```{r}
df <- readRDS(file = paste0(here(), "/data/df_ch-lae.rds"))
```

## Convert ET to mass units

Convert latent heat flux (W/m2) to evapotranspiration in mass units (mm/d).
```{r}
# tested: identical results are obtained with:
# bigleaf::LE.to.ET(LE_F_MDS, TA_F_MDS)* 30 * 30 * 24
le_to_et <- function(le, tc, patm){
  1000 * 30 * 30 * 24 * le / (cwd::calc_enthalpy_vap(tc) * cwd::calc_density_h2o(tc, patm))
}

df <- df |>
  mutate(et = le_to_et(LE_F_MDS, TA_F_MDS, PA_F))

df |> 
  ggplot(aes(et_bigleaf, et)) +
  geom_point()
```

Check annual totals.
```{r}
adf <- df |>
  mutate(year = year(TIMESTAMP)) |>
  group_by(year) |>
  summarise(et = sum(et), prec = sum(P_F))

adf |>
  tidyr::pivot_longer(cols = c(et, prec), names_to = "Flux") |> 
  ggplot(aes(x = year, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)")
```

Each year, ET is greater than precipitation. Hence, the water deficit will not continue accumulating over multiple years.

## Simulate snow

Simulate snow accumulation and melt based on temperature and precipitation.
```{r}
df <- df |>
  mutate(prec = ifelse(TA_F_MDS < 0, 0, P_F),
         snow = ifelse(TA_F_MDS < 0, P_F, 0)) |>
  cwd::simulate_snow(varnam_prec = "prec", varnam_snow = "snow", varnam_temp = "TA_F_MDS")
```

Visualise snow mass equivalent time series.
```{r}
df |> 
  ggplot(aes(TIMESTAMP, snow_pool)) + 
  geom_line() +
  labs(x = "Date", "Snow mass equivalent (mm)")
```

This looks like it's a lot of snow, actually. Maybe the melting rate is too slow.

Define the daily water balance as liquid water infiltrating into soil (taken as rain plus snow melt) minus evapotranspiration - both in mass units, or equivalently in mm/d.
```{r}
df <- df |>
  mutate(wbal = liquid_to_soil - et)
```

Visualise it.
```{r}
gg5 <- df |>
  ggplot(aes(TIMESTAMP, et)) +
  geom_line()

gg6 <- df |>
  ggplot(aes(TIMESTAMP, liquid_to_soil)) +
  geom_line()

gg7 <- df |>
  ggplot(aes(TIMESTAMP, wbal)) +
  geom_line()

gg5 / gg6 / gg7
```

## Cumulative water deficit algorithm

Get CWD and events.
```{r}
out_cwd <- cwd(df,
               varname_wbal = "wbal",
               varname_date = "TIMESTAMP",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)
```

Retain only events of a minimum length of 20 days.
```{r}
out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)
```

Plot CWD time series.
```{r}
ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, prec), size  =  0.3, color = "royalblue") +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 60)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water deficit (mm)")
```

## Extreme value statistics

Get annual maxima and fit a general extreme value distribution using the {extRemes} package.
```{r}
vals <- out_cwd$inst %>%
  group_by(year(date_start)) %>%
  summarise(deficit = max(deficit, na.rm = TRUE)) %>%
  pull(deficit)

evd_gev <- extRemes::fevd(x = vals, type = "GEV", method = "MLE", units = "years")
summary(evd_gev)
```

Get CWD magnitudes for given return periods.
```{r}
return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 200, 250, 300, 500, 800)

return_level <- extRemes::return.level(
  evd_gev,
  return.period = return_period
)
df_return <- tibble(
  return_period = return_period,
  return_level = unname(c(return_level)),
  trans_period = -log( -log(1 - 1/return_period)) )

df_return |> 
  ggplot(aes(return_period, return_level)) +
  geom_point() +
  labs(x = "Return period (yr)", 
       y = "Magnitude of annual CWD maximum (mm)",
       title = "GEV")
```

With a Gumbel extreme value distribution, the return period as a function of the CWD extreme magnitude is calculated as follows:
```{r}
# Fit Gumbel distribution
evd_gumbi <- extRemes::fevd(x = vals, type = "Gumbel", method = "MLE", units = "years")
summary(evd_gumbi)

# calculate return period as a function of the CWD extreme. Using the two 
# coefficients of the fitted distribution as arguments
calc_return_period <- function(x, loc, scale){
  1 / (1 - exp(-exp(-(x-loc)/scale)))
}

extract_loc <- function(mod){
  loc <- mod$results$par[ "location" ]
  if (!is.null(loc)){
    return(loc)
  } else {
    return(NA)
  }
}

extract_scale <- function(mod){
  scale <- mod$results$par[ "scale" ]
  if (!is.null(scale)){
    return(scale)
  } else {
    return(NA)
  }
}

# demo return periods
return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 200, 250, 300, 500, 800)

# use built-in function to get expected CWD extreme for given return periods 
# (inverse of probability)
return_level <- extRemes::return.level(
  evd_gumbi,
  return.period = return_period
)

# create data frame for visualisation
df_return <- tibble(
  return_period = return_period,
  return_level = unname(c(return_level)),
  trans_level = -log( -log(1 - 1/return_period))) |> 
  mutate(myreturn_period = calc_return_period(
    return_level,
    extract_loc(evd_gumbi),
    extract_scale(evd_gumbi)
  ))

# CWD extreme for a given return period
df_return |> 
  ggplot(aes(return_period, return_level)) +
  geom_point() +
  labs(x = "Return period (yr)", 
       y = "Magnitude of annual CWD maximum (mm)",
       title = "Gumbel")

# Return period for a given CWD extreme (calculated based on function above)
df_return |> 
  ggplot(aes(return_level, myreturn_period)) + 
  geom_point() +
  labs(y = "Return period (yr)", 
       x = "Magnitude of annual CWD maximum (mm)",
       title = "Gumbel")
```

Visualise the estimated event size with a return period of $T = 80$ y as the red line on top of the distribution of cumulative water deficit events.
```{r}
ggplot() +
  geom_histogram(
    data = out_cwd$inst,
    aes(x = deficit, y = after_stat(density)),
    color = "black",
    position="identity",
    bins = 6
    ) +
  labs(x = "Cumulative water deficit (mm)") +
  geom_vline(xintercept = df_return %>%
               dplyr::filter(return_period == 80) %>%
               pull(return_level),
             col = "tomato")
```

## Time stepping

The data frame used above contains time series with daily resolution. The CWD algorithm can also be applied to data provided at other time steps. It primarily acts on the *rows* in the data frame.
```{r}
wdf <- df |> 
  mutate(year = lubridate::year(TIMESTAMP),
         week = lubridate::week(TIMESTAMP)) |> 
  group_by(year, week) |> 
  summarise(wbal = sum(wbal, na.rm = FALSE)) |> 
  
  # create a date object again, considering the first day of the week
  mutate(date = lubridate::ymd(paste0(year, "-01-01")) + lubridate::weeks(week-1))

out_cwd_weekly <- cwd(wdf,
               varname_wbal = "wbal",
               varname_date = "date",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)
```

Plot weekly CWD time series.
```{r}
ggplot() +
  geom_rect(
    data = out_cwd_weekly$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_cwd_weekly$df, aes(date, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 60)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water deficit (mm)")
```
