---
title: "Example"
author: "Beni Stocker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{map2tidy functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
```

This demonstrates the workflow for determining cumulative water deficit (CWD) time series and fitting an extreme value distribution to annual maxima of the CWD time series.

## Prepare data

Read data from the file contained in this repository.
```{r}
df <- readRDS(file = paste0(here(), "/data/df_ch-lae.rds"))
```

## Convert ET to mass units

Convert latent heat flux (W/m2) to evapotranspiration in mass units (mm/d).
```{r}
df <- df |>
  mutate(et = cwd::convert_et(LE_F_MDS, TA_F_MDS, PA_F))
```

Check annual totals.
```{r}
adf <- df |>
  mutate(year = year(TIMESTAMP)) |>
  group_by(year) |>
  summarise(et = sum(et), prec = sum(P_F))

adf |>
  tidyr::pivot_longer(cols = c(et, prec), names_to = "Flux") |> 
  ggplot(aes(x = year, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)")
```
Each year, ET is greater than precipitation. Hence, the water deficit will not continue accumulating over multiple years.

## Simulate snow

Simulate snow accumulation and melt based on temperature and precipitation.
```{r}
df <- df |>
  mutate(prec = ifelse(TA_F_MDS < 0, 0, P_F),
         snow = ifelse(TA_F_MDS < 0, P_F, 0)) |>
  cwd::simulate_snow(varnam_prec = "prec", varnam_snow = "snow", varnam_temp = "TA_F_MDS")
```

Visualise snow mass equivalent time series.
```{r}
df |> 
  ggplot(aes(TIMESTAMP, snow_pool)) + 
  geom_line()
```
This looks like it's a lot of snow, actually. Maybe the melting rate is too slow.

Define the daily water balance as liquid water infiltrating into soil (taken as rain plus snow melt) minus evapotranspiration - both in mass units, or equivalently in mm/d.
```{r}
df <- df |>
  mutate(wbal = liquid_to_soil - et)
```

Visualise it.
```{r}
gg5 <- df |>
  ggplot(aes(TIMESTAMP, et)) +
  geom_line()

gg6 <- df |>
  ggplot(aes(TIMESTAMP, liquid_to_soil)) +
  geom_line()

gg7 <- df |>
  ggplot(aes(TIMESTAMP, wbal)) +
  geom_line()

gg5 / gg6 / gg7
```

## Cumulative water deficit algorithm

Get CWD and events.
```{r}
out_cwd <- cwd(df,
               varname_wbal = "wbal",
               varname_date = "TIMESTAMP",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)
```

Retain only events of a minimum length of 20 days.
```{r}
out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)
```

Plot CWD time series.
```{r}
ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, prec), size  =  0.3, color = "royalblue") +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 200)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water deficit (mm)")
```

## Extreme value statistics

Get annual maxima and fit a general extreme value distribution using the {extRemes} package.
```{r}
vals <- out_cwd$inst %>%
  group_by(year(date_start)) %>%
  summarise(deficit = max(deficit, na.rm = TRUE)) %>%
  pull(deficit)

evd_gev <- extRemes::fevd(x = vals, type = "GEV", method = "MLE", units = "years")
summary(evd_gev)
```

Get CWD magnitudes for given return periods.
```{r}
return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 200, 250, 300, 500, 800)

return_level <- extRemes::return.level(
  evd_gev,
  return.period = return_period
)
df_return <- tibble(
  return_period = return_period,
  return_level = unname(c(return_level)),
  trans_period = -log( -log(1 - 1/return_period)) )

df_return |> 
  ggplot(aes(return_period, return_level)) +
  geom_point() +
  labs(x = "Return period (yr)", y = "Magnitude of annual CWD maximum (mm)")
```

Visualise the estimated event size with a return period of $T = 80$ y on top of the distribution of cumulative water deficit events.
```{r}
ggplot() +
  geom_histogram(
    data = out_cwd$inst,
    aes(x = deficit, y = ..density..),
    color = "black",
    position="identity",
    bins = 6
    ) +
  labs(x = "Cumulative water deficit (mm)") +
  geom_vline(xintercept = df_return %>%
               dplyr::filter(return_period == 80) %>%
               pull(return_level),
             col = "tomato")
```
